---
editor_options: 
  markdown: 
    wrap: 72
---

# How R thinks about data 

-   Be able to describe the different data types R uses

-   Use c(), str(), class(), and typeof() to make and investigate
    vectors

-   Understand coercion between data types

-   Know how factors work under the hood

-   Be able to manipulate factors

# Why?

-   R is like speaking another language so its important to understand
    what you are telling R to do

-   learning about how R thinks is fundamental to this

-   really important to know because it will help you trouble shoot
    error message & deal with problems are that are hard to Google

# Vectors

-   most basic way R deals with about data

-   any series of values (numbers or text)

-   we assign these values to an *object*

-   bind values together with c() function

```{r}
weight <- c(50, 60, 65, 82) 
weight

animals <- c("mouse", "rat", "dog") 
animals

# inspection functions
length(weight)
length(animals) 
class(weight_g) 
class(animals) 
str(weight_g) 
str(animals)

# change vectors
weight <- c(weight, 90) # add to the end of the vector 

weight <- c(30, weight) # add to the beginning of the vector weight_g

```

**Challenge**

-   We’ve seen that atomic vectors can be of type character, numeric (or
    double), integer, and logical. But what happens if we try to mix
    these types in a single vector?

-   What will happen in each of these examples? (hint: use `class()` to
    check the data type of your objects):

```{r}
num_char <- c(1, 2, 3, "a")
num_logical <- c(1, 2, 3, TRUE)
char_logical <- c("a", "b", "c", TRUE)
tricky <- c(1, 2, 3, "4")
```

-   Why do you think it happens?

-   How many values in `combined_logical` are `"TRUE"` (as a character)
    in the following example:

```{r}
num_logical <- c(1, 2, 3, TRUE)
char_logical <- c("a", "b", "c", TRUE)
combined_logical <- c(num_logical, char_logical)
```

-   You’ve probably noticed that objects of different types get
    converted into a single, shared type within a vector. In R, we call
    converting objects from one class into another class *coercion*.
    These conversions happen according to a hierarchy, whereby some
    types get preferentially coerced into other types. Can you draw a
    diagram that represents the hierarchy of how these data types are
    coerced?

# Subsetting 

-   If we want to extract one or several values from a vector, we must
    provide one or several indices in square brackets.

    ```{r}
    animals <- c("mouse", "rat", "dog", "cat") animals[2] # could be read as "return the second value in animals"

    animals[c(3, 2)] # could be read as "return the third and second values in animals" weight

    #We can also repeat the indices to create an object with more elements than the original one: 
    animals[c(1, 2, 3, 2, 1, 4)]
    ```

## Conditional subsetting ----

-   Another common way of subsetting is by using a logical vector.
    `TRUE` will select the element with the same index, while `FALSE`
    will not:

-   logical vectors are usually an intermediate step in subsetting

```{r}
weight <- c(21, 34, 39, 54, 55) weight[c(TRUE, FALSE, TRUE, TRUE, FALSE)] # could be read as "give me the first value, not the second value, etc."

weight > 50 # will return logicals with TRUE for the indices that meet the condition

# if we want to actually select values about 50
weight[weight > 50]

#You can combine multiple tests using & (both conditions are true, AND) or | (at least one of the conditions is true, OR): 
weight[weight < 30 | weight > 50]

## different symbols
# >= greater than
# =< less than
# == equal
# %in% within

# Exmple
animals <- c("mouse", "rat", "dog", "cat") animals[animals == "cat" | animals == "rat"] # returns both rat and cat
animals

# VS
animals %in% c("rat", "cat", "dog", "duck", "goat")

## Challenge
# why does this return TRUE?
"four" > "five"
```

## Vector Math 

-   You can add a number to a vector of numbers like this:

```{r}
x <- 1:10 
x + 3 
x * 10

# adding two vectors together of the SAME length
y <- 100:109 
x + y

# different length -- RECYCYLING
z <- 1:2 
x + z
#Whoa… what happened here? R does something called recycling. It adds together the first values of each vector, then the second values, but then it runs out of values in vector z. It then recycles the values of z, so it will add the 3rd value of x to the 1st value of z, then the 4th value of x to the 2nd value of z, and so on. Essentially, it recycles z from 1,2 into 1,2,1,2,1,2,1,2,1,2, then does the addition.

# save as a new object
a <- x + z
# R warns us about this! However, if you try to assign this result to an object, we get the warning, but the assignment works.
```

# Missing data

-   When doing operations on numbers, most functions will return `NA` if
    the data you are working with include missing values.

-   You can add the argument na.rm=TRUE to calculate the result while
    ignoring the missing values

```{r}
heights <- c(2, 4, 4, NA, 6) 
mean(heights) 
mean(heights, na.rm = TRUE)

#Extract those elements which are not missing values.
is.na(heights) # this returns a logical vector with TRUE where there is an NA

!is.na(heights) # the ! means "is not", so now we get a logical vector with FALSE for NAs

heights[!is.na(heights)] # now we put that logical vector in, and it will NOT return the entries with NA

#Extract those elements which are complete cases. The returned object is an atomic vector of type "numeric" (or "double").
heights[complete.cases(heights)]

```

### Challenge

1.  Using this vector of heights in inches, create a new vector with the
    NAs removed.

```{r}
heights <- c(63, 69, 60, 65, NA, 68, 61, 70, 61, 59, 64, 69, 63, 63, NA, 72, 65, 64, 70, 63, 65)
```

2.  Use the function `median()` to calculate the median of the `heights`
    vector.

3.  Use R to figure out how many people in the set are taller than 67
    inches.

# Other Data Structures

Vectors are one of the many **data structures** that R uses. Other
important ones are lists (`list`), data frames (`data.frame`), matrices
(`matrix`), arrays (`array`), and factors (`factor`). These are all
built from combinations of vectors, so much of what you learned about
vectors will be important when working with these data structures.

-   lists: multiple vectors, dif data types

-   data frame: most common, basically just a picky list that has to
    have the same length, multiple data types

-   matrices and arrays: single type of data, not super common

-   factors: fancier vector

    # Why factors

-   Factors can be convenient at times, and they will pop up pretty
    frequently, but in most circumstances, character strings will give
    you fewer hassles.

-   It’s usually best to start with character vectors, and convert them
    explicitly to factors if you need to.

-   Some functions in R will automatically convert character strings to
    factors. For instance, read.csv() run in older versions R will turn
    any character data into factors, while in newer versions this has
    been changed to keep them as characters.

-   If you aren’t sure, you can use the argument stringsAsFactors=FALSE
    in read.csv() to make sure your character strings as character
    strings.

    ```{r}
    # Factors are used to represent categorical data
    animals <- factor(c("duck", "duck", "goose", "goose")) 
    class(animals) 
    typeof(animals) 
    levels(animals) 
    nlevels(animals) 

    ## current order 
    animals <- factor(animals, levels = c("goose", "duck")) animals # after re-ordering
    ```

## convert factors

```{r}
as.character(animals) # returns index values of the characters 
# When is this an issue? 
year_fct <- factor(c(1990, 1983, 1977, 1998, 1990)) #WRONG 
as.numeric(year_fct)
as.numeric(as.character(year_fct)) # This does the trick
```

### Challenge

-   Copy, paste and run the code below in your R script:\
    `treatment <- factor(c("high", "low", "low", "medium", "high"))`

-   First, re-order the levels of `treatment` so that “low” is first,
    “medium” is second, and “high” is third. Hint: Use the `factor()`
    function again, but with an additional `levels` argument.

-   Next, check the names with the `levels()` function, then use this
    same function to rename the levels of treatment to “L”, “M” and “H”

```{r}
treatment <- factor(c("high", "low", "low", "medium", "high"))
treatment <- factor(treatment, levels = c("low", "mediam", "high"))
levels(treatment) <- c("L", "M", "H")
treatment


```
